Bow shocks, bow waves, and dust waves. II. Beyond the rip point

### 2019-03-07 papers/Papers1234/sec-strong-coupling.tex
Finish first complete draft of Paper I

We now consider the application to bow shocks around main sequence OB
stars, as well as cool and hot supergiants, expressing stellar and
ambient parameters in terms of typical values as follows:
\begin{align*}
  \label{eq:stellar-parameters}
  \dot{M}_{-7} &= \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 &= V / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 &= L / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} &= v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n &= (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} &= \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{align*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).  Note that \(\kappa = \SI{600}{cm^2.g^{-1}}\) corresponds to a
cross section of \(\approx \SI{e-21}{cm^2}\) per hydrogen nucleon, which is
typical for interstellar medium dust \citep{Bertoldi:1996a} at far
ultraviolet wavelengths, where OB stars emit most of their radiation.
In terms of these parameters, we can express the stellar wind momentum
efficiency as
\begin{equation}
  \label{eq:wind-eta-typical}
  \eta\wind = \num{0.495} \,\dot{M}_{-7} \,V_3  \,L_4^{-1}
\end{equation}
and the fiducial radius and optical depth as
\begin{align}
  \label{eq:Rstar-typical}
  R_* / \si{pc} &= \num{2.21} \, (L_4 / n)^{1/2} \,v_{10}^{-1} \\
  \label{eq:taustar-typical}
  \tau_* &= \num{0.0089} \,\kappa_{600} \, (L_4 \,n)^{1/2} \,v_{10}^{-1} \ .
\end{align}
In Figure~\ref{fig:zones-v-n-plane}, we show results for the bow size
(apex distance, \(R_0\)) as a function of the density, \(n\), and
relative velocity, \(v_\infty\), of the external stream, with each panel
corresponding to a particular star, with parameters as shown in
Table~\ref{tab:stars}.  To facilitate comparison with previous work,
we choose stellar parameters similar to those used in the
hydrodynamical simulations of \citet{Meyer:2014b, Meyer:2016a,
  Meyer:2017a}, based on stellar evolution tracks for stars of
\SIlist{10;20;40}{M_\odot} \citep{Brott:2011a} and theoretical wind
prescriptions \citep{de-Jager:1988a, Vink:2000a}.  Although the
stellar parameters do evolve with time, they change relatively little
during the main-sequence lifetime of several million years.\footnote{%
  \label{fn:meyer-velocities-too-low}
  Note that we have recalculated the stellar wind terminal velocities,
  since the values given in the \citeauthor{Meyer:2014b} papers are
  troublingly low.  We have used the prescription
  \(V = 2.6 V_{\text{esc}}\), where
  \(V_{\text{esc}} = \left( 2 G M (1 - \Gamma_e)/ R \right)^{1/2}\) is the
  photospheric escape velocity, which is appropriate for strong
  line-driven winds with \(T_{\text{eff}} > \SI{21 000}{K}\)
  \citep{Lamers:1995a}.  We find velocities of
  \SIrange{2500}{3300}{km.s^{-1}}, which are consistent with
  observations and theory \citep{Vink:1999a}, but at least two times
  higher than those cited by \citet{Meyer:2014b}. } %
The three examples are an early B~star (\SI{10}{M_\odot}), a late O~star
(\SI{20}{M_\odot}), and an early O~star (\SI{40}{M_\odot}), which cover the
range of luminosities and wind strengths expected from bow-producing
hot main sequence stars.  The luminosity is a steep function of
stellar mass (\(L \sim M^{2.5}\)) and the wind mass-loss rate is a steep
function of luminosity (\(\dot{M} \sim L^{2.2}\)), which means that the
wind momentum efficiency is also a steep function of mass
(\(\eta\wind \sim M^3\)), approaching unity for early O~stars, but falling
to less than 1\% for B~stars.

@@

\subsection{Environments of bow-forming stars}
\label{sec:popul-bow-prod}

@@

In \S~\ref{sec:depend-stell-type} we found that the bow is
wind-supported when \(\tau_*^2 < \eta\wind\), where \(\tau_*\) is a fiducial
optical depth (eq.~[\ref{eq:taustar-typical}]) and \(\eta\wind\) is the
radiative momentum efficiency of the stellar wind
(eq.~[\ref{eq:wind-eta-typical}]).  This translates into a maximum
stream density for wind-supported bows of
\newcommand\crit{\ensuremath{_{\text{max}}}}
\newcommand\vink{\ensuremath{_{\text{\tiny Vink}}}}
\begin{equation}
  \label{eq:ncrit}
  n\crit \approx \left(\SI{30}{cm^{-3}}\right)
  \left(  \frac{v} {\SI{10}{km.s^{-1}}}\right)^2
  \left( \frac{\eta\wind}{\eta\vink} \right)
  \left( \frac{\SI{600}{cm^2.g^{-1}}}{\kappa} \right)
  \ ,
\end{equation}
which is largely insensitive to spectral type between early-B and
early-O stars.  In this equation, we have expressed the wind
efficiency in terms of the Table~\ref{tab:stars} values,
\(\eta\vink\), which are calculated from the \citet{Vink:1999a} recipe,
and the UV dust opacity in terms of the standard ISM value adopted in
\S~\ref{sec:depend-stell-type}.

@@

For cases where the stellar wind or dust properties differ from our
assumed values, then these conclusions may change.  For instance,
there are a growing number of stars, for which very low mass-loss
rates have been diagnosed from UV, optical or infrared line profiles
\citep{Martins:2005b, Marcolino:2009a, Najarro:2011a, Martins:2012a,
  Smith:2017b}.  Most of these stars are late-type O~dwarfs, for which
the \citet{Vink:2000a} prescription predicts a mass-loss rate of
\num{3e-8} to \SI{e-7}{M_\odot.yr^{-1}}, whereas the observed values are
in the range \num{e-10} to \SI{3e-9}{M_\odot.yr^{-1}} -- a shortfall of 10
to 1000!  This ``weak wind problem'' is a far larger discrepancy than
can be explained by clumping effects.  A potential (partial)
resolution is to suppose that internal shocks near the base of the
wind, together with thermal conduction, heat a large fraction of the
gas to coronal temperatures, where it is no longer detectable by means
of standard wind diagnostic lines \citep{Lucy:2012a}.  However, in
cases where x-ray diagnostics have been used to trace the mass loss
this hot component (\(\mu\)~Col, \citealp{Huenemoerder:2012a};
\(\zeta\)~Oph, \citealp{Cohen:2014a}), rates that are still an order of
magnitude below the Vink values are found.  We therefore show in

\subsection{Impact of the weak-wind phenomenon}
\label{sec:impact-weak-wind}

@@

Hot stars, comparison with expected velocities and densities.

Expected velocities.  There are cases of high velocities, e.g., AE~Aur, but they are probably rare. \citep{Renzo:2018a}
In cases, where proper motions are measured, they are often low.

Metallicity dependence of wind strength. RBW more likely at lower metallicities.

Weak wind stars - repeat the 20 Msun case, but for a much lower \(\eta\wind\)

@@

Cool stars, comparison with observed densities (Betelgeuse, for
instance). Other ones in \citep{Cox:2012a}.

Prospects for observation in LMC: 1 parsec = 4 arcsec, so need JWST to get most of them.

\subsection{Diagnosing bow type from shell emission profile}
\label{sec:diagnosing-bow-type}

Given that the predicted shell density profile varies between
wind-supported and radiation-supported bows
(\S~\ref{sec:axial-structure-bow} and Fig.~\ref{fig:axial-structure}),
it is worth considering whether it may be possible to use observed
emission profiles to distinguish between the two.

>>>>

We now consider the application to bow shocks around main sequence OB
stars, as well as cool and hot supergiants, expressing stellar and
ambient parameters in terms of typical values as follows:
\begin{align*}
  \label{eq:stellar-parameters}
  \dot{M}_{-7} &= \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 &= V / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 &= L / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} &= v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n &= (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} &= \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{align*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).  Note that \(\kappa = \SI{600}{cm^2.g^{-1}}\) corresponds to a
cross section of \(\approx \SI{e-21}{cm^2}\) per hydrogen nucleon, which is
typical for interstellar medium dust \citep{Bertoldi:1996a} at far
ultraviolet wavelengths, where OB stars emit most of their radiation.
In terms of these parameters, we can express the stellar wind momentum
efficiency as
\begin{equation}
  \label{eq:wind-eta-typical}
  \eta\wind = \num{0.495} \,\dot{M}_{-7} \,V_3  \,L_4^{-1}
\end{equation}
and the fiducial radius and optical depth as
\begin{align}
  \label{eq:Rstar-typical}
  R_* / \si{pc} &= \num{2.21} \, (L_4 / n)^{1/2} \,v_{10}^{-1} \\
  \label{eq:taustar-typical}
  \tau_* &= \num{0.0089} \,\kappa_{600} \, (L_4 \,n)^{1/2} \,v_{10}^{-1} \ .
\end{align}
In Figure~\ref{fig:zones-v-n-plane}, we show results for the bow size
(apex distance, \(R_0\)) as a function of the density, \(n\), and
relative velocity, \(v_\infty\), of the external stream, with each panel
corresponding to a particular star, with parameters as shown in
Table~\ref{tab:stars}.  To facilitate comparison with previous work,
we choose stellar parameters similar to those used in the
hydrodynamical simulations of \citet{Meyer:2014b, Meyer:2016a,
  Meyer:2017a}, based on stellar evolution tracks for stars of
\SIlist{10;20;40}{M_\odot} \citep{Brott:2011a} and theoretical wind
prescriptions \citep{de-Jager:1988a, Vink:2000a}.  Although the
stellar parameters do evolve with time, they change relatively little
during the main-sequence lifetime of several million years.\footnote{%
  \label{fn:meyer-velocities-too-low}
  Note that we have recalculated the stellar wind terminal velocities,
  since the values given in the \citeauthor{Meyer:2014b} papers are
  troublingly low.  We have used the prescription
  \(V = 2.6 V_{\text{esc}}\), where
  \(V_{\text{esc}} = \left( 2 G M (1 - \Gamma_e)/ R \right)^{1/2}\) is the
  photospheric escape velocity, which is appropriate for strong
  line-driven winds with \(T_{\text{eff}} > \SI{21 000}{K}\)
  \citep{Lamers:1995a}.  We find velocities of
  \SIrange{2500}{3300}{km.s^{-1}}, which are consistent with
  observations and theory \citep{Vink:2000a}, but at least two times
  higher than those cited by \citet{Meyer:2014b}. } %
The three examples are an early B~star (\SI{10}{M_\odot}), a late O~star
(\SI{20}{M_\odot}), and an early O~star (\SI{40}{M_\odot}), which cover the
range of luminosities and wind strengths expected from bow-producing
hot main sequence stars.  The luminosity is a steep function of
stellar mass (\(L \sim M^{2.5}\)) and the wind mass-loss rate is a steep
function of luminosity (\(\dot{M} \sim L^{2.2}\)), which means that the
wind momentum efficiency is also a steep function of mass
(\(\eta\wind \sim M^3\)), approaching unity for early O~stars, but falling
to less than 1\% for B~stars.

@@

@@

In \S~\ref{sec:depend-stell-type} we found that the bow is
wind-supported when \(\tau_*^2 < \eta\wind\), where \(\tau_*\) is a fiducial
optical depth (eq.~[\ref{eq:taustar-typical}]) and \(\eta\wind\) is the
radiative momentum efficiency of the stellar wind
(eq.~[\ref{eq:wind-eta-typical}]).  This translates into a maximum
stream density for wind-supported bows of
\newcommand\crit{\ensuremath{_{\text{max}}}}
\newcommand\vink{\ensuremath{_{\text{\tiny Vink}}}}
\begin{equation}
  \label{eq:ncrit}
  n\crit \approx \left(\SI{30}{cm^{-3}}\right)
  \left(  \frac{v} {\SI{10}{km.s^{-1}}}\right)^2
  \left( \frac{\eta\wind}{\eta\vink} \right)
  \left( \frac{\SI{600}{cm^2.g^{-1}}}{\kappa} \right)
  \ ,
\end{equation}
which is largely insensitive to spectral type between early-B and
early-O stars.  In this equation, we have expressed the wind
efficiency in terms of the Table~\ref{tab:stars} values,
\(\eta\vink\), which are calculated from the \citet{Vink:2000a} recipe,
and the UV dust opacity in terms of the standard ISM value adopted in
\S~\ref{sec:depend-stell-type}.

@@

@@

For cases where the stellar wind or dust properties differ from our
assumed values, then these conclusions may change.  For instance,
there are a growing number of stars, for which very low mass-loss
rates have been diagnosed from UV, optical or infrared line profiles
\citep{Martins:2005b, Marcolino:2009a, Najarro:2011a, Martins:2012a,
  Shenar:2017a, Smith:2017b}.  Most of these stars are late-type
O~dwarfs, for which the \citet{Vink:2000a} prescription predicts a
mass-loss rate of \num{3e-8} to \SI{e-7}{M_\odot.yr^{-1}}, whereas the
observed values are in the range \num{e-10} to \SI{3e-9}{M_\odot.yr^{-1}}
-- a shortfall of 10 to 1000!  This ``weak wind problem'' is a far
larger discrepancy than can be explained by clumping effects.  A
potential (partial) resolution is to suppose that internal shocks near
the base of the wind, together with thermal conduction, heat a large
fraction of the gas to coronal temperatures, where it is no longer
detectable by means of standard wind diagnostic lines
\citep{Lucy:2012a}.  However, in cases where x-ray diagnostics have
been used to trace the mass loss of this hot component (\(\mu\)~Col,
\citealp{Huenemoerder:2012a}; \(\zeta\)~Oph, \citealp{Cohen:2014a}), rates
that are still an order of magnitude below the Vink values are found
(see also \citealp{Shenar:2017a}).  In Figure~\ref{fig:O-weak-wind},
we show results for a weak-wind O9 dwarf, which is identical to the
\SI{20}{M_\odot} star from Table~\ref{tab:stars}, except with
\(\eta\wind = 0.01 \eta\vink\) (approximating the measurements from the
sample of \citealp{Martins:2005b}).  For such stars, the boundary
between the wind-supported and radiation-supported regimes shifts to
lower ambient densities, reaching \(n\crit \sim \SI{1}{cm^{-3}}\) for
\(v = \SI{20}{km.s^{-1}}\).  This means that radiation-supported bows
become feasible even in the diffuse ISM for this important class of
stars (if winds are really as weak as the UV diagnostics suggest).  It
is therefore important to critically re-evaluate attempts to
\emph{measure} mass loss rates using observations of bows
\citep{Gvaramadze:2012a, Kobulnicky:2018a} since such calculations
invariably assume that the WBS regime is universally applicable.  This
issue is addressed in detail in Paper~III.

Variations in the UV dust opacity would also effect our results.
These may be due to variations in the dust/gas ratio or to changes in
the size distribution of the grains.  The UV opacity is dominated by
very small grains (VSG) with radius \(a \sim \SI{10}{nm}\), which
represent a small fraction (\(\simeq 5\%\)) of the total dust mass in
standard grain mixtures \citep{Desert:1990a}.  This same VSG
population dominates the mid-infrared continuum emission around
\SI{24}{\um}, which is where stellar bows are most easily detected
\citep{Meyer:2016a, Kobulnicky:2016a}. There is some evidence that
such grains may be depleted in the ionized gas of the Orion Nebula
\citep{Salgado:2016a}, leading to a decrease in \(\kappa\) by a factor of
about six.  By equation~\eqref{eq:ncrit}, this would increase
\(n\crit\) by the same factor, making radiation-supported bows less
likely.  On the other hand, the process of Radiative Torque Disruption
is predicted to be important \citep{Hoang:2018a} for the high
radiation fields found in stellar bows, and this would tend to
\emph{increase} the VSG abundance (and hence \(\kappa\)) via the
centrifugal destruction of larger grains.  In LMC \hii{} regions,
\citet{Stephens:2014b} find observational evidence that the VSG
abundance increases as the radiation field gets stronger.

Variations in metallicity will also be reflected in
\(\kappa\), assuming that the total dust-gas ratio is roughly proportional
to the metal abundance, \(Z\).  However, the wind strength
\(\eta\wind\) will also increases with metallicity
(\(\dot{M} V\wind \propto Z^{0.82}\), \citealp{Vink:2001a}), which will
largely cancel out the effect in our bow models.

@@

% Cool stars, comparison with observed densities (Betelgeuse, for
% instance). Other ones in \citep{Cox:2012a}.

% Prospects for observation in LMC: 1 parsec = 4 arcsec, so need JWST to get most of them.

\section{Summary}
\label{sec:summary}

We present a systematic study of the formation of emission arcs, or
bows, around luminous stars that move supersonically relative to their
surrounding medium, taking into account both the stellar wind and
radiation pressure of the star.  In this initial study, we consider
the case where gas and grains are perfectly coupled via collisions,
and apply our models to OB stars.  Our principal results are as
follows:
\begin{enumerate}[1.]
\item Three different regimes of interaction are possible, in order of
  increasing optical depth of the bow shell: Wind-supported Bow Shock
  (WBS), Radiation-supported Bow Wave (RBW), and Radiation-supported
  Bow Shock (RBS).
\item For a broad range of stellar types, the WBS regime occurs when
  the ambient density is below a critical density:
  \(n\crit \approx \SI{100}{cm^{-3}}\) for a typical relative velocity of
  \SI{20}{km.s^{-1}}.
\item The critical density increases for faster moving stars, but
  decreases for stars with weak stellar winds.
\item At densities higher than \(n\crit\), B~stars tend to be in the
  RBW regime, and O~stars in the RBS regime.
\item For main sequence OB stars, the bow shell remains fully
  photoionized for ambient densities up to about 100 times
  \(n\crit\). For isolated B~supergiants, on the other hand, the
  ionization front may be trapped by the shell even in the WBS regime.
\item The bow shell in the RBW regime tends to be broad, with the
  density gradually ramping up at the inner and outer edges, whereas
  in the WBS and RBS regimes the shell is thinner with more sharply
  defined edges.
\item Studies that have estimated wind mass-loss rates from
  observations of bows need to be re-evaluated in the light of
  possible radiation support. 
\end{enumerate}
